{\small} CUBEP3M: Closer to optimal N-body computations

### 2012-04-13 Introduction.tex
Started to work on the Poissonsolver.tex and Structure.tex

PMFAST was developed in XXX \cite{ref:PMFAST} and was already highly efficient, although the resolution was limited by the absence of particle-particle interaction at sub-grid level. Also, the force was obtained by solving Poisson's equation for the the gravitational potential, which led to the force by finite difference. 
Finally, the simulation volume was decomposed into a set of slabs, which becomes problematic when we try to parallelize the calculation.
Each computing unit is assigned a sub-volume of the slab, which eventually reach the limit where one cannot divide the finest cell grid, in a non-adaptative mesh grid at least.

>>>>

PMFAST was developed in XXX \cite{ref:PMFAST} (we refer to this paper as MPT hereafter) and was already highly efficient, although the resolution was limited by the absence of particle-particle interaction at sub-grid level. Also, the force was obtained by solving Poisson's equation for the the gravitational potential, which led to the force by finite difference. 
Finally, the simulation volume was decomposed into a set of slabs, which becomes problematic when we try to parallelize the calculation.
Each computing unit is assigned a sub-volume of the slab, which eventually reach the limit where one cannot divide the finest cell grid, in a non-adaptative mesh grid at least.

### 2012-04-13 PoissonSolver.tex
Started to work on the Poissonsolver.tex and Structure.tex

This section describes how Poisson's equation is solved
within {\small CUBEP3M}.

It is of course possible to shut down any of these components,

Double mesh levels, equation to match them, fudge factor, kernel, 
corrected kernel, pp and extended pp, fft to solve the force.

>>>>

This section describes how Poisson's equation is solved at each time step of the main loop. 
Many parts of the algorithm are identical to {\small PMFAST}, hence we refer extensively 
to section 2 of MPT.

The reason that motivates the double mesh configuration is to allow the force of gravity to be broken down into 
a long and a short range components. This two-mesh decomposition was first seen in Hydra \citep{hydra}, 
and is now used by many codes, including {\small PMFAST}, {\bf (TracPen2003)} and {\bf (Other codes?)}.
The long range force requires knowledge of the complete simulated volume, 
but can be computed on a coarser grid without degrading significantly the precision of the calculation.
Using a grid 64 times coarser limits the global communication time, plus contribute only to a fraction of the total memory requirement.
The short range force is computed on the finer mesh and is local in the sense that it only knows about the matter in the immediate surroundings.
It can thus be computed simultaneously on many computing units with {\small OPENMP} parallel loops, each {\small CPU} solving for a different sub-volume,
as long as the arrays are constructed such as to support parallel reading and writing\footnote{In practice, this is done by adding an additional dimension
to the relevant arrays, such that each {\small CPU} accesses a unique memory location.}. 
The particle-particle (pp) interactions are also local operations, and are computed inside the same {\small OPENMP} loops.

The force of gravity on a mesh can be computed either with a gravitational potential  kernel $\omega_{\phi} ({\bf x})$
  or a force  kernel $\omega_{F} ({\bf x})$.
It is a curl-free field, which allows us to relate the potential $\phi({\bf x})$to the source term via Poisson's equation: 
\begin{eqnarray}
\nabla^{2}\phi({\bf x}) = 4 \pi G \rho({\bf x})
\label{eq:poisson}
\end{eqnarray}
$G$ being Newton's constant. We solve this equation in Fourier space, where we write
\begin{eqnarray}
 \tilde{\phi}({\bf k}) = \frac{4 \pi G \tilde{\rho}({\bf k})}{- k^{2}} \equiv \tilde{\omega}_{\phi}({\bf k})\tilde{\rho}({\bf k})
\label{eq:poissonFourier}
\end{eqnarray}
The potential in real space is then obtained with an inverse Fourier transform, and the kernel becomes $\omega_{\phi} ({\bf x}) = -G/r$.
Using the convolution theorem, we can write
\begin{eqnarray}
 \phi({\bf x}) = \int \rho({\bf x'}) \omega_{\phi}({\bf x'} - {\bf x}) d{\bf x'}
\label{eq:poisson_solution_pot}
\end{eqnarray}
Although this approach is fast, it involves a finite differentiation which enhances the numerical noise.
We therefore opt for a force kernel, which is more accurate but has the inconvenient to require four extra Fourier transforms.
In this case, we must solve the convolution in three dimensions:
\begin{eqnarray}
 F({\bf x}) = - m {\bf \nabla} \phi({\bf x})   = \int \rho({\bf x'}) {\bf  \omega}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
\label{eq:poisson_solution_force}
\end{eqnarray}
The differentiation does not affect the density since it only acts on un-prime variables,
and the force kernel is given by 
\begin{eqnarray}
{\bf  \omega}_{F}({\bf x}) \equiv - {\bf \nabla}\omega_{\phi}({\bf x}) = - \frac{mG \hat{\bf r}}{r^{2}}
\end{eqnarray}

Following MPT, we split  the force kernel into two components, for the short and long range respectively, and 
match the overlapping regions with a polynomial. Namely, we have:
\begin{eqnarray}
{\bf \omega}_{s}(r) = \begin{cases} {\bf \omega}_{F}(r) -  {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
0 & \mbox{otherwise} 
\end{cases}
\end{eqnarray}
and
\begin{eqnarray}
{\bf \omega}_{l}(r) = \begin{cases} {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
 {\bf\omega}_{F}(r)  &\mbox{otherwise} 
\end{cases}
\end{eqnarray}
The vector $ {\bf \beta}(r)$ is related to the fourth order polynomial that is used in the potential case by
 $ {\bf \beta} = - {\bf \nabla} \alpha(r)$. The coefficients are found by matching the boundary conditions at $r_{c}$ up to the second derivative,
 and we get
  \begin{eqnarray}
   {\bf \beta}(r) = \bigg[ -\frac{7 r}{4 r_{c}^{3}} + \frac{3 r^{3}}{4 r^{5}}\bigg] \hat{\bf r}
  \end{eqnarray}
The long range force is subject to a correction that fixes {\bf (what exactly?)}.

 equation to match them, fudge factor,
corrected kernel, pp and extended pp.

### 2012-04-13 Structure.tex
Started to work on the Poissonsolver.tex and Structure.tex

The code flow is presented in Fig. \ref{fig:structure}.
Before entering the main loop, the code starts with an initialization stage, 
in which many declared variables are assigned default values,
the redshift checkpoints are read, the {\small FFTW} plans are created, and the {\small MPI} communicators are defined.
The phase-space array  is obtained from the output of the initial condition generator,
and the force kernels on both grids are also constructed from the specific geometry of the simulation.
For clarity, these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
although they are actually distinct calls in the code.

Each iteration of the main loop starts with the {\tt timestep} subroutine, 
which proceed to a determination of the redshift jump by comparing the step size constraints from each
force components and from the cosmic expansion.
The latter is found by Taylor expanding  Friedmann's equation up to the third order in the scale factor.
Poisson's equation is then solved  in the {\tt particle\_mesh} subroutine,
which updates the positions and velocities of the dark matter particles with a leapfrog scheme \citep{Hockney}.
If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
and write them to file.

@@

The {\tt particle\_mesh} subroutine is conceptually identical to that of {\small PMFAST}, with the exception  that {\small CUBEP3M}
decomposes the volume into cubes (as opposed to slabs). The loop over tiles is thus performed in three dimension.
A schematic representation of the subroutine is shown in Fig. \ref{fig:particle_mesh}, 
and we save the detailed description of the force calculation  in section \ref{sec:Poisson}.

\begin{figure}
\begin{verbatim}
subroutine particle_mesh
   call update_position
   call particle_pass
   call link_list
   !$omp parallel do
   do tile = 1, tiles_node
      call rho_f_ngp
      call cmplx_rho_f
      call kernel_multiply_f
      call force_f
      call update_velocity_f      
      call link_list_pp
      call force_pp
      call update_velocity_pp
   end do
   !$omp end parallel do
   call rho_c_ngp
   call cmplx_rho_c
   call kernel_multiply_c
   call force_c
   call update_velocity_c      
   delete_buffers
end subroutine particle_mesh
\end{verbatim}
\caption{Overall structure of the two-level mesh algorithm.}
\label{fig:particle_mesh}
\end{figure}

force calculation,

>>>>

The code flow is presented in Fig. \ref{fig:structure}.
Before entering the main loop, the code starts with an initialization stage, 
in which many declared variables are assigned default values,
the redshift checkpoints are read, the {\small FFTW} plans are created, and the {\small MPI} communicators are defined.
The phase-space array  is obtained from the output of the initial condition generator,
and the force kernels on both grids are also constructed from the specific geometry of the simulation.
These kernels are slightly modified in the code configuration where the exact Newtonian force is extended outside a fine grid cell.
For clarity, all these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
although they are actually distinct calls in the code.

Each iteration of the main loop starts with the {\tt timestep} subroutine, 
which proceed to a determination of the redshift jump by comparing the step size constraints from each
force components and from the cosmic expansion.
The latter is found by Taylor expanding  Friedmann's equation up to the third order in the scale factor.
Poisson's equation is then solved  in the {\tt particle\_mesh} subroutine,
which updates the positions and velocities of the dark matter particles with a leapfrog scheme \citep{Hockney} {\bf (more on leapfrog?)}.
If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
and write them to file.

@@

The {\tt particle\_mesh} subroutine is conceptually identical to that of {\small PMFAST}, with the exception  that {\small CUBEP3M}
decomposes the volume into cubes (as opposed to slabs). The loop over tiles is thus performed in three dimension.
A schematic representation of the subroutine is shown in Fig. \ref{fig:particle_mesh}.
We save the detailed description of the force calculation  in section \ref{sec:Poisson}.

\begin{figure}
\begin{verbatim}
subroutine particle_mesh
   call update_position
   call particle_pass
   call link_list
   !$omp parallel do
   do tile = 1, tiles_node
      call rho_f_ngp
      call cmplx_rho_f
      call kernel_multiply_f
      call force_f
      call update_velocity_f      
      call link_list_pp
      call force_pp
      call update_velocity_pp
      if(extended_pp = .true.)
         call link_list_pp_extended
         call force_pp_extended
         call update_velocity_pp_extended       
      endif
   end do
   !$omp end parallel do
   call rho_c_ngp
   call cmplx_rho_c
   call kernel_multiply_c
   call force_c
   call update_velocity_c      
   delete_buffers
end subroutine particle_mesh
\end{verbatim}
\caption{Overall structure of the two-level mesh algorithm.}
\label{fig:particle_mesh}
\end{figure}
